---
title: "Mini-Project #1"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Libraries

```{r}
library(tidyverse)
library(jsonlite)
```

# Data

```{r}
apple_raw <- read_csv("mini_proj1/apple_earnings.csv")

apple <- apple_raw |>
  mutate(paras = map(TRANSCRIPTS, fromJSON)) |>
  select(-TRANSCRIPTS) |>
  unnest(paras)

apple
```
```{r}
glimpse(apple)
```



# Introduction

An earnings call is a quarterly conference call held by companies to share their financial performance with investors and the public. Earnings calls are often used by investors to fundamentally analyze the financial health of a company and forecast future performance. These calls typically include an introduction, disclaimers, reports from the CEO and CFO, as well as open Q&A from external analysts. 

Our dataset, taken from fruitstanddev on Kaggle, provides transcripts of Apple earnings calls from 2005-2025. Each earnings call is split into paragraphs, each labeled by speaker. Through out sentiment analysis, we seek to examine how tone shifts around company-level and macroeconomic events as well as the relationship between sentiment and stock performance.


# Data Cleaning

## Eliminate Empty Content Lines

```{r}
apple2 <- apple |>
  filter(str_detect(content, "[a-zA-Z]"))
```

## Standardize Speaker Names

```{r}
apple2 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

```{r}
apple3 <- apple2 |>
  mutate(
    speaker = str_remove(
      speaker,
      "^[QA]\\s*[-\u2013]\\s*|\\s+[-\u2013].*$|,.*$"
    ),
    speaker = str_remove(
      speaker,
      "\\s[a-zA-Z].?(?=\\s)"
    ),
    speaker = str_replace(
      speaker,
      "^([A-Z][a-z]+)([A-Z])",
      "\\1 \\2"
    ),
    speaker = str_remove_all(
      speaker,
      "\\s[A-Z][a-z]+(?=\\s)"
    ),
    
  )
```

```{r}
apple3 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```



```{r}
apple3 <- apple2 |>
  mutate(standardized_speaker = case_when(
    is.na(speaker) ~ "None",
    str_detect(speaker, "Nancy Paxton") ~ "Nancy Paxton",
    str_detect(speaker, "Peter Oppenheimer") ~ "Peter Oppenheimer",
    str_detect(speaker, "Tejas Gala") ~ "Tejas Gala",
    str_detect(speaker, "Tim.*Cook") ~ "Tim Cook",
    str_detect(speaker, "Luca.*Maestri") ~ "Luca Maestri",
    str_detect(speaker, "Sacconaghi") ~ "Antonio Sacconaghi",
    str_detect(speaker, "Aaron.*Rakers") ~ "Aaron Rakers",
    str_detect(speaker, "Amit") ~ "Amit Daryanani",
    str_detect(speaker, "Andrew Neff") ~ "Andrew Neff",
    str_detect(speaker, "Andy Hargreaves") ~ "Andy Hargreaves",
    str_detect(speaker, "Atif Malik") ~ "Atif Malik",
    str_detect(speaker, "Ben.*Reitzes") ~ "Ben Reitzes",
    str_detect(speaker, "Ben Bollin") ~ "Ben Bollin",
    str_detect(speaker, "Shope") ~ "Bill Shope",
    str_detect(speaker, "Bill Fearnley") ~ "Bill Fearnley",
    TRUE ~ speaker
  ))
```


```{r}
apple3 |>
  group_by(standardized_speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

