---
title: "Mini-Project #1"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false

# load libraries

library(tidyverse)
library(jsonlite)

# read data into correct format

apple_raw <- read_csv("mini_proj1/apple_earnings.csv")

apple <- apple_raw |>
  mutate(paras = map(TRANSCRIPTS, fromJSON)) |>
  select(-TRANSCRIPTS) |>
  unnest(paras)

apple
```

# Introduction

An earnings call is a quarterly conference call held by companies to share their financial performance with investors and the public. Earnings calls are often used by investors to fundamentally analyze the financial health of a company and forecast future performance. These calls typically include an introduction, disclaimers, reports from the CEO and CFO, as well as open Q&A from external analysts. 

Our dataset, taken from fruitstanddev on Kaggle, provides transcripts of Apple earnings calls from 2005-2025. Each earnings call is split into paragraphs, each labeled by speaker. Through our sentiment analysis, we seek to examine how tone shifts around company-level and macroeconomic events as well as the relationship between sentiment and stock performance.

```{r}
#| echo: false

# use str_detect to eliminate empty content lines

apple2 <- apple |>
  filter(str_detect(content, "[a-zA-Z]"))
```

```{r}
#| echo: false

# check speaker naming conventions

apple2 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

```{r}
#| echo: false

# use str_remove and str_detect to standardize speaker names

apple3 <- apple2 |>
  mutate(
    speaker = str_remove(
      speaker,
      "^[QA]\\s*[-\u2013]\\s*|\\s+[-\u2013].*$|,.*$"
    ),
    speaker = str_remove(
      speaker,
      "\\s[a-zA-Z].?(?=\\s)"
    ),
    speaker = str_replace(
      speaker,
      "^([A-Z][a-z]+)([A-Z])",
      "\\1 \\2"
    ),
    speaker = str_remove_all(
      speaker,
      "\\s[A-Z][a-z]+(?=\\s)"
    ),
    speaker = case_when(
        str_detect(speaker, "Cook")        ~ "Tim Cook",
        str_detect(speaker, "Jobs")        ~ "Steve Jobs",
        str_detect(speaker, "Sacconaghi")  ~ "Toni Sacconaghi",
        str_detect(speaker, "Reitzes")     ~ "Ben Reitzes",
        str_detect(speaker, "Bollin")      ~ "Ben Bollin",
        str_detect(speaker, "Shope")       ~ "Bill Shope",
        str_detect(speaker, "Marshall")    ~ "Brian Marshall",
        str_detect(speaker, "Munster")     ~ "Gene Munster",
        str_detect(speaker, "Caso")        ~ "Chris Caso",
        str_detect(speaker, "Whit.*more")  ~ "Chris Whitmore",
        str_detect(speaker, "Suva")        ~ "Jim Suva",
        str_detect(speaker, "Huberty")     ~ "Katy Huberty",
        str_detect(speaker, "^Keith B")    ~ "Keith Bachman",
        str_detect(speaker, "Sankar")      ~ "Krish Sankar",
        str_detect(speaker, "Abramsky")    ~ "Mike Abramsky",
        str_detect(speaker, "\\sNg$")      ~ "Mike Ng",
        str_detect(speaker, "Olson")       ~ "Mike Olson",
        str_detect(speaker, "Gardi?ner")   ~ "Richard Gardner",
        str_detect(speaker, "Semple")      ~ "Rob Semple",
        str_detect(speaker, "\\sHo$")      ~ "Sidney Ho",
        str_detect(speaker, "Milunovich")  ~ "Steve Milunovich",
        str_detect(speaker, "^Gary W")     ~ "Gary Wipfler",
        TRUE ~ speaker
      )
    )
```

### Check standardized names

```{r}
#| echo: false

# check speaker naming standardization

apple3 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

```{r}
#| echo: false

# filter mislabeled or empty rows

apple4 <- apple3 |>
  filter(!(speaker %in% c("Executives", "Analysts", "NA", "Unknown", "Apple Computer", "Piper Jaffray", "Apple's 2005 Call"))) |>
  filter(!(is.na(speaker))) |>
  filter(!(REPORT_DATE %in% as.Date(c("2005-10-13", "2007-04-25")))) |>
  mutate(speaker = if_else(speaker == "", "Unknown Analyst", speaker))
```

```{r}
#| echo: false

# label rows as executive speaker, operations, or analyst

executives <- c("Tim Cook", "Steve Jobs", "Peter Oppenheimer", "Luca Maestri", "Kevan Parekh")
operations <- c("Nancy Paxton", "Tejas Gala", "Saori Casey", "Suhasini Chandramouli", "Operator")

apple5 <- apple4 |>
  mutate(role = case_when(
    speaker %in% executives ~ "Executive",
    speaker %in% c(operations) ~ "Operations",
    TRUE ~ "Analyst"
  ))
```

```{r}
#| echo: false

# label data as Q&A or prepared

apple_final <- apple5 |>
  group_by(REPORT_DATE) |>
  mutate(
    qa_start = min(paragraph_number[role == "Analyst"]),
    section = if_else(paragraph_number >= qa_start, "Q&A", "Prepared")
  ) |>
  ungroup()
```




