---
title: "Mini-Project #1"
format:
  pdf: default
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| echo: false

# load libraries

library(tidyverse)
library(jsonlite)
library(tidytext)

# read data into correct format

apple_raw <- read_csv("mini_proj1/apple_earnings.csv")

apple <- apple_raw |>
  mutate(paras = map(TRANSCRIPTS, fromJSON)) |>
  select(-TRANSCRIPTS) |>
  unnest(paras)

apple
```

# Introduction

An earnings call is a quarterly conference call held by companies to share their financial performance with investors and the public. Earnings calls are often used by investors to fundamentally analyze the financial health of a company and forecast future performance. These calls typically include an introduction, disclaimers, reports from the CEO and CFO, as well as open Q&A from external analysts. 

Our dataset, taken from fruitstanddev on Kaggle, provides transcripts of Apple earnings calls from 2005-2025. Each earnings call is split into paragraphs, each labeled by speaker. Through our sentiment analysis, we seek to examine how tone shifts around company-level and macroeconomic events as well as the relationship between sentiment and stock performance.

```{r}
#| echo: false

# use str_detect to eliminate empty content lines

apple2 <- apple |>
  filter(str_detect(content, "[a-zA-Z]"))
```

```{r}
#| echo: false

# check speaker naming conventions

apple2 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

```{r}
#| echo: false

# use str_remove and str_detect to standardize speaker names

apple3 <- apple2 |>
  mutate(
    speaker = str_remove(
      speaker,
      "^[QA]\\s*[-\u2013]\\s*|\\s+[-\u2013].*$|,.*$"
    ),
    speaker = str_remove(
      speaker,
      "\\s[a-zA-Z].?(?=\\s)"
    ),
    speaker = str_replace(
      speaker,
      "^([A-Z][a-z]+)([A-Z])",
      "\\1 \\2"
    ),
    speaker = str_remove_all(
      speaker,
      "\\s[A-Z][a-z]+(?=\\s)"
    ),
    speaker = case_when(
        str_detect(speaker, "Cook")        ~ "Tim Cook",
        str_detect(speaker, "Jobs")        ~ "Steve Jobs",
        str_detect(speaker, "Sacconaghi")  ~ "Toni Sacconaghi",
        str_detect(speaker, "Reitzes")     ~ "Ben Reitzes",
        str_detect(speaker, "Bollin")      ~ "Ben Bollin",
        str_detect(speaker, "Shope")       ~ "Bill Shope",
        str_detect(speaker, "Marshall")    ~ "Brian Marshall",
        str_detect(speaker, "Munster")     ~ "Gene Munster",
        str_detect(speaker, "Caso")        ~ "Chris Caso",
        str_detect(speaker, "Whit.*more")  ~ "Chris Whitmore",
        str_detect(speaker, "Suva")        ~ "Jim Suva",
        str_detect(speaker, "Huberty")     ~ "Katy Huberty",
        str_detect(speaker, "^Keith B")    ~ "Keith Bachman",
        str_detect(speaker, "Sankar")      ~ "Krish Sankar",
        str_detect(speaker, "Abramsky")    ~ "Mike Abramsky",
        str_detect(speaker, "\\sNg$")      ~ "Mike Ng",
        str_detect(speaker, "Olson")       ~ "Mike Olson",
        str_detect(speaker, "Gardi?ner")   ~ "Richard Gardner",
        str_detect(speaker, "Semple")      ~ "Rob Semple",
        str_detect(speaker, "\\sHo$")      ~ "Sidney Ho",
        str_detect(speaker, "Milunovich")  ~ "Steve Milunovich",
        str_detect(speaker, "^Gary W")     ~ "Gary Wipfler",
        TRUE ~ speaker
      )
    )
```

### Check standardized names

```{r}
#| echo: false

# check speaker naming standardization

apple3 |>
  group_by(speaker) |>
  summarize(
    n= n()
  ) |>
  print(n = Inf)
```

```{r}
#| echo: false

# filter mislabeled or empty rows

apple4 <- apple3 |>
  filter(!(speaker %in% c("Executives", "Analysts", "NA", "Unknown", "Apple Computer", "Piper Jaffray", "Apple's 2005 Call"))) |>
  filter(!(is.na(speaker))) |>
  filter(!(REPORT_DATE %in% as.Date(c("2005-10-13", "2007-04-25")))) |>
  mutate(speaker = if_else(speaker == "", "Unknown Analyst", speaker))
```

```{r}
#| echo: false

# label rows as executive speaker, operations, or analyst

executives <- c("Tim Cook", "Steve Jobs", "Peter Oppenheimer", "Luca Maestri", "Kevan Parekh")
operations <- c("Nancy Paxton", "Tejas Gala", "Saori Casey", "Suhasini Chandramouli", "Operator")

apple5 <- apple4 |>
  mutate(role = case_when(
    speaker %in% executives ~ "Executive",
    speaker %in% c(operations) ~ "Operations",
    TRUE ~ "Analyst"
  ))
```

```{r}
#| echo: false

# label data as Q&A or prepared

apple_final <- apple5 |>
  group_by(REPORT_DATE) |>
  mutate(
    qa_start = min(paragraph_number[role == "Analyst"]),
    section = if_else(paragraph_number >= qa_start, "Q&A", "Prepared")
  ) |>
  ungroup()
```

# Sentiment Over Time

```{r}
apple_final$
```



```{r}
loughran <- get_sentiments("loughran")

apple_exec_words <- apple_final |>
  filter(role == "Executive") |>
  unnest_tokens(word, content, token = "words") |>
  group_by(REPORT_DATE, section) |>
  mutate(total_words = n()) |>
  ungroup()

apple_exec_words |>
  inner_join(loughran, relationship = "many-to-many") |>
  count(REPORT_DATE, sentiment, total_words, section) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(net = ((positive - negative) / total_words) * 1000) |>
  ggplot(aes(x = REPORT_DATE, y = net, fill = net > 0)) +
      geom_col() + 
      facet_wrap(~section)
```


```{r}
loughran <- get_sentiments("loughran")

apple_exec_words <- apple_final |>
  filter(role == "Executive") |>
  unnest_tokens(word, content, token = "words") |>
  group_by(REPORT_DATE, section) |>
  mutate(total_words = n()) |>
  ungroup()

apple_exec_words |>
  inner_join(loughran, relationship = "many-to-many") |>
  count(REPORT_DATE, sentiment, total_words, section) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |> 
  mutate(net = ((positive - negative) / total_words) * 100) |>
  ggplot(aes(x = REPORT_DATE, y = net)) +
      geom_col(fill = "#0071E3") + 
      facet_wrap(~section) +
      theme_minimal() + 
      labs(
        title = "Net Loughran Sentiment for Apple Executives per 100 Words",
        x = "Earnings Call Date",
        y = "Net Sentiment per 100 Words",
      )
```

This is a bar chart showing the net Loughran sentiment for Apple executives per 100 words. Loughran sentiment is a finance-specific sentiment lexicon, and net sentiment per 100 words is calculated by subtracting the number of negative words from the number of positive words, and then dividing by the total number of words and multiplying by 100 to standardize for the length of each earnings call. The graph is then faceted into sentiment in the prepared remarks from executives and the Q&A executive statements. Earnings call dates span from 2006 to 2025. It is also important to know that rows were dropped from 2005, 2007, and 2013, due to missing data. Net sentiment per 100 words ranges from 0 to 3.

Sentiment remains net positive across all earnings calls in both the prepared and Q&A sections. For the prepared section, sentiment trends in the positive direction over time. Q&A sentiment remains generally stable, on average, despite fluctuations. Sentiment is significantly higher across time for prepared remarks in comparison to Q&A remarks.